---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.height = 5,
  fig.width = 5
#  out.width = "100%"
)

library(genridge)
```

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/105555707.svg)](https://zenodo.org/badge/latestdoi/105555707)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/genridge)](https://cran.r-project.org/package=genridge)
[![](http://cranlogs.r-pkg.org/badges/grand-total/genridge)](https://cran.r-project.org/package=genridge)
<!-- badges: end -->

# genridge

## Generalized Ridge Trace Plots for Ridge Regression

Version 0.6-8

### Overview

The genridge package introduces generalizations of the standard univariate
ridge trace plot used in ridge regression and related methods (Friendly, 2013)  These graphical methods
show both bias (actually, shrinkage) and precision, by plotting the covariance ellipsoids of the estimated
coefficients, rather than just the estimates themselves.  2D and 3D plotting methods are provided,
both in the space of the predictor variables and in the transformed space of the PCA/SVD of the
predictors.  

### Details

This package provides computational support for the graphical methods described in Friendly (2013). Ridge regression models may be fit using the function `ridge`, which incorporates features of `MASS::lm.ridge` and `ElemStatLearn::simple.ridge`. In particular, the shrinkage factors in ridge regression may be specified either in terms of the constant ($\lambda$) added to the diagonal of $X^T X$ matrix, or the equivalent number of degrees of freedom.

More importantly, the `ridge` function also calculates and returns the associated covariance matrices of each of the ridge estimates, allowing precision to be studied and displayed graphically.

This provides the support for the main plotting functions in the package:

* `plot.ridge`: Bivariate ridge trace plots, showing the covariance ellipse of the estimated coefficients.

* `pairs.ridge`: All pairwise bivariate ridge trace plots

* `plot3d.ridge`: 3D ridge trace plots

* `traceplot`: Traditional univariate ridge trace plots

In addition, the function `pca.ridge` transforms the coefficients and covariance matrices of a ridge object from predictor space to the equivalent, but more interesting space of the PCA of $X^T X$ or the SVD of X. The main plotting functions also work for these objects, of class `c("ridge", "pcaridge")`.

Finally, the functions `precision` and `vif.ridge` provide other useful measures and plots.

## Installation

Get the _released_ version from CRAN:

     install.packages("genridge")

The _development_ version can be installed to your R library directly from this repo via:

     if (!require(devtools)) install.packages("devtools")
     library(devtools)
     install_github("friendly/genridge")

## Examples

The classic example for ridge regression is Longley's (1967) data, consisting of 7 economic variables, observed yearly from 1947 to 1962 (n=16), in the data frame `datasets::longley`.
The goal is to predict `Employed` from `GNP`, `Unemployed`, `Armed.Forces`, `Population`, `Year`,
`GNP.deflator`.

These data are perverse, in that 
* each variable is a time series so that there is clearly a lack of independence
* there is also some _structural colinearity_ among the variables `GNP`, `Year`, `GNP.deflator`, `Population`.

```{r longley1}
data(longley)
longley.y <- longley[, "Employed"]
longley.X <- data.matrix(longley[, c(2:6,1)])
```

Shrinkage values, can be specified using either `\lambda` (where $\lambda = 0$ corresponds to ordinary
least squares (OLS)),
or equivalent effective degrees of freedom.
This quantifies the tradeoff between bias and variance for predictive modeling, 
where OLS has low bias, but can have large predictive variance.

`ridge` returns a matrix containing the coefficients for each predictor for each shrinkage value.
```{r longley2}
lambda <- c(0, 0.005, 0.01, 0.02, 0.04, 0.08)
lridge <- ridge(longley.y, longley.X, lambda=lambda)
lridge
```

Do the same, using formula interface:
```{r longley3}
lridge <- ridge(Employed ~ GNP + Unemployed + Armed.Forces + Population + Year + GNP.deflator, 
		data=longley, lambda=lambda)
```

## Variance Inflation Factors

`vif()` for a `"ridge"` object calculates variance inflation factors for all
values of the ridge constant. You can see that for OLS ($\lambda = 0$), nearly all VIF values
are dangeroulsly high. With a ridge factor of 0.04 or greater, variance inflation has been
considerably reduced.

```{r vif}
vif(lridge)
```


### Univariate trace plots

A standard, univariate, trace plot simply plots the estimated coefficients for each predictor
against the shrinkage factor, $\lambda$.
```{r longley-tp1}
traceplot(lridge)
```

It is sometimes easier to interpret the plot when coefficients are plotted against the equivalent
degrees of freedom, where $\lambda = 0$ corresponds to 6 degrees of freedom in the parameter
space of six predictors.
```{r longley-tp2}
traceplot(lridge, X="df")
```

These plots show the trends in increased bias associated with larger $\lambda$, but they do not
show the accompaning decrease in variance. For that, we need to consider the variances and
covariances of the estimated coefficients.

### Bivariate trace plots

The bivariate analog of the trace plot suggested by Friendly (2013) plots bivariate
confidence ellipses for pairs of coefficients. Here, we plot those for `GNP` against
four of the other predictors.

```{r longley-plot-ridge}
#| out.width = "100%",
#| fig.show = "hold"
op <- par(mfrow=c(2,2), mar=c(4, 4, 1, 1)+ 0.1)
clr <-  c("black", "red", "darkgreen","blue", "cyan4", "magenta")
pch <- c(15:18, 7, 9)
lambdaf <- c(expression(~widehat(beta)^OLS), ".005", ".01", ".02", ".04", ".08")

for (i in 2:5) {
	plot.ridge(lridge, variables=c(1,i), 
	           radius=0.5, cex.lab=1.5, col=clr, 
	           labels=NULL, fill=TRUE, fill.alpha=0.2)
	text(lridge$coef[1,1], lridge$coef[1,i], 
	     expression(~widehat(beta)^OLS), cex=1.5, pos=4, offset=.1)
	text(lridge$coef[-1,c(1,i)], lambdaf[-1], pos=3, cex=1.3)
}
par(op)
```
The `pairs` method for `ridge` objects shows this in scatterplot matrix form.

```{r longley-pairs}
#| out.width = "100%"
pairs(lridge, radius=0.5)
```


## Low-rank views

Just as principal components analysis gives low-dimensional views of a data set, PCA can
be useful to understand ridge regression.

The function `pca.ridge` transforms a `ridge` object
from parameter space, where the estimated coefficients are
$\beta_k$ with covariance matrices $\Sigma_k$, to the
principal component space defined by the right singular vectors, $V$,
of the singular value decomposition of the scaled predictor matrix, $X$.

```{r pca-traceplot}
plridge <- pca.ridge(lridge)
traceplot(plridge)
```

what is perhaps surprising is that the coefficients for the first 4 components are not shrunk at all.
Rather, the effect of ridge regression is seen only on the last two dimensions.

```{r pca-pairs}
#| out.width = "100%"
pairs(plridge)
```

```{r pca-dim56}
plot(plridge, variables=5:6, fill = TRUE, fill.alpha=0.2)
```


## References

Friendly, M. (2013).
The Generalized Ridge Trace Plot: Visualizing Bias *and* Precision.
*Journal of Computational and Graphical Statistics*, **22**(1), 50-68,
[doi link](http://dx.doi.org/10.1080/10618600.2012.681237), Preprint:
[genridge.pdf](http://euclid.psych.yorku.ca/datavis/papers/genridge.pdf)

Longley, J. W.  (1967) An appraisal of least-squares programs from the point of view of the user. 
_Journal of the American Statistical Association_,
**62**, 819–841.
